---
title: "Coursework - Data Science II"
author: "Omar Zhadykov, 220220503"
output:
  html_notebook:
    fig_width: 10
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document: default
  html_document:
    fig_width: 10
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
---

<script>
$(document).ready(function() {
  $items = $('div#TOC li');
  $items.each(function(idx) {
    num_ul = $(this).parentsUntil('#TOC').length;
    $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
  });

});
</script>

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}
library(svglite)
library(knitr)
suppressPackageStartupMessages(library(data.table))
library(ggplot2)
knitr::opts_chunk$set(dev = "svglite")

# Put your dataset in the same folder as your R file. This code will set your working directory for this notebook to the folder where the R file is stored. This way I can rerun your code without modifications.

library(rstudioapi)
setwd(dirname(getActiveDocumentContext()$path))
```

# Introduction

In this course work I will explore dataset about students' performance based on their habits and lifestyle. I have chosen this dataset, because it is rich on different discrete, continious variables with a good range and I assume good variance, which will be very useful in Data Science 2 class. 

This dataset consists of 16 columns which could be categorized into these groups:
  - discrete variables (example: exercise frequency per week)
  - continious variables (example: sleep quantity)
  - categorical variables (example par time job, yes or no)
  
This is regression problem, because I will try to predict student's score for an exam, based on different variables.

# Description

### Collection

Here I would like to collect, prepare, and explore my data. First thing is to import the data set.

```{r}
dt_students <- fread(file = "./data/student_habits_performance.csv")
```

<br>

I would like to check, if i have some nullish data in my dataset. I think it is a good idea to go through all rows and colums and check, if there is a NA. I want to check it with built-in function in R *complete.cases(data_table)*. This function returns TRUE or FALSE if row contains a NA value.

```{r}
nas <- dt_students[!complete.cases(dt_students)]
nas
```

That looks great. Now we can move on to exploration. But before I start, It is crucial to install all needed libraries.

```{r}
library(data.table)
library(ggcorrplot)
library(ggExtra)
library(ggplot2)
library(skimr)
library(ggridges)
library(ggsci)
library(ggthemes)
library(RColorBrewer)
library(svglite)
library(viridis)
library(scales)
library(rpart)
library(rpart.plot)
library(factoextra)
```

### Exploration

I found some helpful functions in R, so we could have a look on our data. We will start with a structure, than we will get some statistic data and take a *head()* of the data

```{r}
str(dt_students)
```
<br>
Statistic data:
```{r}
skim(dt_students)
```

So we have 1000 rows, 16 columns, 7 char values and 9 numeric values. This is great start to get an overall opinion, so let's move.

<br>
This is a sample of dataset:

```{r}
head(dt_students)
```

I would like to start from density of a main values, which are from my domain knowledge are important for the best performance at the university.

Attendance percentage density: 

```{r}
ggplot(data = dt_students, aes(x = attendance_percentage)) + 
  geom_density(fill="#f1b147", color="#f1b147", alpha=0.25) + 
  labs(
    x = 'Attendance',
    y = 'Density'
  ) +
  geom_vline(xintercept = mean(dt_students$attendance_percentage), linetype="dashed") + 
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```
This density plot visualizes the distribution of the attendance percentage, showing that most students attend classes at a rate of ~85% roughly, and this is a right-skewed distribution. The dashed vertical line represents the mean attendance percentage (~84-85%). The plot shows, that the majority of students are attending most of the classes.


Age density:

```{r}
ggplot(data = dt_students, aes(x = age)) + 
  geom_density(fill="#f1b147", color="#f1b147", alpha=0.25) + 
  labs(
    x = 'Age',
    y = 'Density'
  ) +
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

Age density plot looks pretty well distributed, so I think there is a better chance to learn good model from this dataset and generalize well.


<br>

Next plot will visualize the distribution of exam scores depending on attendance, and I would like to take a look at how possibly hours studied per day affect the exam score. 

```{r}
ggplot() + 
  geom_point(data = dt_students, aes(x = attendance_percentage, y = exam_score, color = study_hours_per_day)) +
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

This scatter plot visualizes the relationship between attendance percentage (x-axis) and exam score (y-axis), with color indicating the amount of hours studied per day. It shows a positive correlation between high attendance, more study hours per day and exam score. I would say that on average students, who learn and attend classes more tend to get better scores. However, there is some variability, as some students with attendance more than 85% and average studying hours get worst scores.

The next plot, which I am going to do is a boxplot and I want to use diet quality as a factor variable on x axis and exam score on y-axis, to get an overall understanding, how the diet quality may affect the exam score.

```{r}
ggplot(data = dt_students, aes(x = factor(diet_quality), y = exam_score)) +
  geom_boxplot() + 
  theme_minimal() 
```

Boxplot shows that diet quality could be not a very valuable predictor variable. But, we can see, that poor diet has a little bit lower performance. From my personal opinion this difference is so statistically low, that it will not play a big role, but this is still interesting, that If person wants to get the best performance, it is also important to get "the best fuel".

Another very important thing is sleep. I would like to know if there is visible relation between sleep, exercise frequency and exam score, and also sleep distribution of the students.

```{r}
ggplot(data = dt_students, aes(x = sleep_hours, y = exam_score, colour = exercise_frequency)) +
  geom_point() +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000")) +
  scale_colour_gradient(low = "green", high = "darkgreen")
```

```{r}
ggplot(data = dt_students, aes(x = sleep_hours)) + 
  geom_density(fill="#f1b147", color="#f1b147", alpha=0.25) + 
  labs(
    x = 'Sleep',
    y = 'Density'
  ) +
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

From this two plots it is clear that it is not very clear that sleep and sport play a very important role, because there are students who sleep, does sports and gets great scores, but also students, who does not sleep as much, and doesn't so much of a sport, but still gets good grades. Even some Students with 4.5 hours and 1 workaout a week are getting 100% exam score.


Let's create one more boxplot to look at average notes of students with and without part time jobs:

```{r}
ggplot(data = dt_students, aes(x = factor(part_time_job), y = exam_score)) +
  geom_boxplot() + 
  theme_minimal() 
```

Boxplot shows some little difference, but as i thought it is roughly the same.


Now I would like to take a look at one more scatter plot with amount of social media hours with exam score and gender.

```{r}
ggplot(data = dt_students, aes(x = social_media_hours, y = exam_score, colour = gender)) +
  geom_point() +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000"))
```

I would say that I expected another result, but it seems, that on average amount of social media usage does not play a very big role, but it is also an slight average better to spend less time on social media to get better scores.

One last plot will include mental health rating, which from my point of view is very important to perform well.

```{r}
ggplot(data = dt_students, aes(x = mental_health_rating, y = exam_score, colour = gender)) +
  geom_point() +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000"))
```

This plot does not provide enough information, so let's take a look on a vertical barchart.

```{r}
ggplot(data = dt_students, aes(x = mental_health_rating)) + 
  geom_histogram(fill="#2f9e44", color="#2f9e44", alpha=0.25) + 
  geom_vline(xintercept = mean(dt_students$mental_health_rating), linetype="dashed") + 
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```


```{r}
mean(dt_students$mental_health_rating)
```

Okay, from the last 2 plots I could conclude, that mental health could possibly better results on average, but students can still perform well, while being in a "bad" mood.


I think that would be enough exploration and we can start with models.


# Models 1 & 2
- Evaluate and compare your models based on a reasonable evaluation metric of your choice. You must use the same metric for both models. Report both the training and the CV loss. 

First, I would like to start pretty simple with linear model.

I consider to take all variables to my model, because they all seem to be very important.

## Linear model

I will use lm function in R to find needed beta coefficients and create my model

```{r}
price_lm <- lm(formula = price ~ area + bedrooms + hotwaterheating + airconditioning + stories + mainroad + parking + furnishingstatus + bathrooms + guestroom + basement + prefarea, data = dt_houses)

summary(price_lm)
```

We got 0.68 R-squared, which is not that bad for a model just made up. But that's not all, I will try to do better here, but first, another model.

But I would like to measure performance of my models with RMSE, so I will calculate RMSE for linear model.

```{r}
price_lm_rmse <- mean(sqrt(abs(price_lm$residuals)))

price_lm_rmse
```

## Tree Model

I think this model could perform better, because there some variables which can affect this model not only linearly, but the other way, in this case tree model can show better performance.

In this coursework will be used rpart to create a regression tree.

```{r}
prices_tree <- rpart(data = dt_houses, formula = price ~ area + bedrooms + hotwaterheating + airconditioning + stories + mainroad + parking + furnishingstatus + bathrooms + guestroom + basement + prefarea, method = 'anova')

prp(prices_tree, digits = -3)
```

```{r}
printcp(prices_tree)
```

Now after I have built with the help of rpart tree model based on my dataset, let us explore it:

```{r}
prices_tree
```

We can see, that we have 31 Nodes, I think for this kind of dataset it may be okay.

Now it would be great to prune the tree, because I do not want my tree to overfit:

```{r}
plotcp(prices_tree)
```
This is complexity of this tree. We need the lowest complexity, to get as few leafs as possible to get the best performance, so that tree won't overfit the data.

```{r}
prices_tree_min_cp <- prices_tree$cptable[which.min(prices_tree$cptable[, "xerror"]), "CP"]
model_tree <- prune(prices_tree, cp = prices_tree_min_cp )
prp(prices_tree,digits = -3)
```

after we pruned the tree, let's calculate the RMSE for the tree model


```{r}
prices_tree_pred <- predict(prices_tree, dt_houses[, c("area","bathrooms", "bedrooms", "hotwaterheating", "airconditioning", "parking", "stories", "mainroad", "furnishingstatus", "guestroom", "basement", "prefarea")])
prices_tree_rmse <- mean(sqrt(abs(dt_houses$price - prices_tree_pred)))

prices_tree_rmse
```

# Ensemble 

- Repeat the analysis one ensemble method of your choice. 
- Investigate the hyperparameter settings of your ensemble with regards to your evaluation metric. 
- Report both the training and the CV loss. 
- Select the best configurations of your ensemble model based on the same evaluation metric as before.

# Neural Network

- Repeat the analysis with a neural network. 
- Investigate three different configurations with regards to your evaluation metric and select the best configuration. Use the same evaluation metric as before.
- Report both the training and the CV loss. 

# Model Comparison

- Compare the performance of the 4 models.

# PCA

- Run a PCA on your input variables and discuss the scope for dimensionality reduction in your dataset.
- Rerun the previous 4 models on all PCs or on a reduced number of PCs.

### Calculating PCs

Before I start working with PCA, this is important to normalize the data, so that measurement scale will not affect PCs.

```{r}
dt_pca <- data.table(scale(dt_houses[,c("area", "bedrooms", "bathrooms", "stories", "parking")]))
```

Now I want to run my PCA with help of prcomp and get the summary, to dive in to Data.

```{r}
dt_houses_pca <- prcomp(dt_pca)
summary(dt_houses_pca)
```
That looks interesting, but I would like to plot it, to visualize it, so that it could be easier to understand.

```{r}
fviz_eig(dt_houses_pca, addlabels = TRUE) + 
         theme(plot.title = element_text(hjust = 0.5))
```

From this plot it is obvious that all PC are useful and reducing dimensionality will not be beneficial, because last 3 PCs contribute approx. 12% each, which is a fair amount in this case. So I would like to use all PCs, because it is not always necessary to cut dimensionality and it depends on how much variablse we actually have, and how big of a contribution makes each PC. May be in this case "change of basis" will perform better.

### Running models with PCs

Now as I have calculated PCs, it is time to run models with new inputs.

#### Linear model

First we will start off with linear model.

```{r}
pc_table <- dt_houses_pca$x

dt_houses_pc_table_for_lm <- data.frame(price = dt_houses$price, pc_table)

price_lm_pca <- lm(formula = price ~ ., data = dt_houses_pc_table_for_lm)

summary(price_lm_pca)
```

```{r}
price_lm_pca_rmse <- mean(sqrt(abs(price_lm_pca$residuals)))

price_lm_pca_rmse
```


Now, after PCA the results are worse and I think this is because there are fair amount of binary variables in this dataset and PCA could not capture all information, espesially information, which is captured in binary variables, such as "mainroad", "airconditioning" and so on.

 
# Model Selection

- Compare all of your models based on their CV performance. You will have 8 options to consider: four models with and without PCA.
- Present the results in a table or chart.
- Estimate the expected loss of your best model on the test set.

***








