---
title: "Coursework - Data Science II"
author: "Omar Zhadykov, 220220503"
output:
  html_notebook:
    fig_width: 10
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document: default
  html_document:
    fig_width: 10
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options: 
  markdown: 
    wrap: 72
---

<script>
$(document).ready(function() {
  $items = $('div#TOC li');
  $items.each(function(idx) {
    num_ul = $(this).parentsUntil('#TOC').length;
    $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
  });

});
</script>

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}
library(svglite)
library(knitr)
suppressPackageStartupMessages(library(data.table))
library(ggplot2)
knitr::opts_chunk$set(dev = "svglite")

# Put your dataset in the same folder as your R file. This code will set your working directory for this notebook to the folder where the R file is stored. This way I can rerun your code without modifications.

#library(rstudioapi)
#setwd(dirname(getActiveDocumentContext()$path))

library(here)
setwd(here())

print_results <- function(train, cv, test) {
  cat("Training MSE: ", train, "\n")
  cat("CV Loss: ", cv, "\n")
  cat("Test MSE: ", test)
}

```

# Introduction

In this course work I will explore dataset about students' performance
based on their habits and lifestyle. I have chosen this dataset, because
it is rich on different discrete, continious variables with a good range
and I assume good variance, which will be very useful in Data Science 2
class.

This dataset consists of 16 columns which could be categorized into
these groups: - discrete variables (example: exercise frequency per
week) - continious variables (example: sleep quantity) - categorical
variables (example par time job, yes or no)

This is regression problem, because I will try to predict student's
score for an exam, based on different variables.

### Collection

Here I would like to collect, prepare, and explore my data. First thing
is to import the data set.

```{r}
dt_students <- fread(file = "./data/student_habits_performance.csv")
```

<br>

I would like to check, if i have some nullish data in my dataset. I
think it is a good idea to go through all rows and colums and check, if
there is a NA. I want to check it with built-in function in R
*complete.cases(data_table)*. This function returns TRUE or FALSE if row
contains a NA value.

```{r}
nas <- dt_students[!complete.cases(dt_students)]
nas
```

That looks great. Now we can move on to exploration. But before I start,
It is crucial to install all needed libraries and create helper function, to make all models replicable.

```{r}
library(data.table)
library(ggcorrplot)
library(ggExtra)
library(ggplot2)
library(skimr)
library(ggridges)
library(ggsci)
library(ggthemes)
library(RColorBrewer)
library(caret)
library(svglite)
library(viridis)
library(scales)
library(rpart)
library(recipes)
library(rpart.plot)
library(factoextra)
library(keras3)
library(tensorflow)
library(rsample)
library(dplyr)
library(reticulate)

set_seed_for_random_state <- function () {
  # this is for replication
  set.seed(123)
  tensorflow::tf$random$set_seed(123)
  py_run_string("import numpy as np; import random; random.seed(123); np.random.seed(123)")
}
```

### Exploration

I found some helpful functions in R, so we could have a look on our
data. We will start with a structure, than we will get some statistic
data and take a *head()* of the data

```{r}
str(dt_students)
```

<br> Statistic data:

```{r}
skim(dt_students)
```

So we have 1000 rows, 16 columns, 7 char values and 9 numeric values.
This is great start to get an overall opinion, so let's move.

<br> This is a sample of dataset:

```{r}
head(dt_students)
```

I would like to start from density of a main values, which are from my
domain knowledge are important for the best performance at the
university.

Attendance percentage density:

```{r}
ggplot(data = dt_students, aes(x = attendance_percentage)) + 
  geom_density(fill="#f1b147", color="#f1b147", alpha=0.25) + 
  labs(
    x = 'Attendance',
    y = 'Density'
  ) +
  geom_vline(xintercept = mean(dt_students$attendance_percentage), linetype="dashed") + 
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

This density plot visualizes the distribution of the attendance
percentage, showing that most students attend classes at a rate of \~85%
roughly, and this is a right-skewed distribution. The dashed vertical
line represents the mean attendance percentage (\~84-85%). The plot
shows, that the majority of students are attending most of the classes.

Age density:

```{r}
ggplot(data = dt_students, aes(x = age)) + 
  geom_density(fill="#f1b147", color="#f1b147", alpha=0.25) + 
  labs(
    x = 'Age',
    y = 'Density'
  ) +
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

Age density plot looks pretty well distributed, so I think there is a
better chance to learn good model from this dataset and generalize well.

<br>

Next plot will visualize the distribution of exam scores depending on
attendance, and I would like to take a look at how possibly hours
studied per day affect the exam score.

```{r}
ggplot() + 
  geom_point(data = dt_students, aes(x = attendance_percentage, y = exam_score, color = study_hours_per_day)) +
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

This scatter plot visualizes the relationship between attendance
percentage (x-axis) and exam score (y-axis), with color indicating the
amount of hours studied per day. It shows a positive correlation between
high attendance, more study hours per day and exam score. I would say
that on average students, who learn and attend classes more tend to get
better scores. However, there is some variability, as some students with
attendance more than 85% and average studying hours get worst scores.

The next plot, which I am going to do is a boxplot and I want to use
diet quality as a factor variable on x axis and exam score on y-axis, to
get an overall understanding, how the diet quality may affect the exam
score.

```{r}
ggplot(data = dt_students, aes(x = factor(diet_quality), y = exam_score)) +
  geom_boxplot() + 
  theme_minimal() 
```

Boxplot shows that diet quality could be not a very valuable predictor
variable. But, we can see, that poor diet has a little bit lower
performance. From my personal opinion this difference is so
statistically low, that it will not play a big role, but this is still
interesting, that If person wants to get the best performance, it is
also important to get "the best fuel".

Another very important thing is sleep. I would like to know if there is
visible relation between sleep, exercise frequency and exam score, and
also sleep distribution of the students.

```{r}
ggplot(data = dt_students, aes(x = sleep_hours, y = exam_score, colour = exercise_frequency)) +
  geom_point() +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000")) +
  scale_colour_gradient(low = "green", high = "darkgreen")
```

```{r}
ggplot(data = dt_students, aes(x = sleep_hours)) + 
  geom_density(fill="#f1b147", color="#f1b147", alpha=0.25) + 
  labs(
    x = 'Sleep',
    y = 'Density'
  ) +
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

From this two plots it is clear that it is not very clear that sleep and
sport play a very important role, because there are students who sleep,
does sports and gets great scores, but also students, who does not sleep
as much, and doesn't so much of a sport, but still gets good grades.
Even some Students with 4.5 hours and 1 workaout a week are getting 100%
exam score.

Let's create one more boxplot to look at average notes of students with
and without part time jobs:

```{r}
ggplot(data = dt_students, aes(x = factor(part_time_job), y = exam_score)) +
  geom_boxplot() + 
  theme_minimal() 
```

Boxplot shows some little difference, but as i thought it is roughly the
same.

Now I would like to take a look at one more scatter plot with amount of
social media hours with exam score and gender.

```{r}
ggplot( data = dt_students, aes(x = social_media_hours, y = exam_score, colour = gender)) +
  geom_point() +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000"))
```

I would say that I expected another result, but it seems, that on
average amount of social media usage does not play a very big role, but
it is also an slight average better to spend less time on social media
to get better scores.

One last plot will include mental health rating, which from my point of
view is very important to perform well.

```{r}
ggplot(data = dt_students, aes(x = mental_health_rating, y = exam_score, colour = gender)) +
  geom_point() +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000"))
```

This plot does not provide enough information, so let's take a look on a
vertical barchart.

```{r}
ggplot(data = dt_students, aes(x = mental_health_rating)) + 
  geom_histogram(fill="#2f9e44", color="#2f9e44", alpha=0.25) + 
  geom_vline(xintercept = mean(dt_students$mental_health_rating), linetype="dashed") + 
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

```{r}
mean(dt_students$mental_health_rating)
```

Okay, from the last 2 plots I could conclude, that mental health could
possibly better results on average, but students can still perform well,
while being in a "bad" mood.

I think that would be enough exploration and we can start with models.

# Models 1 & 2

Now the interesting part! So before I start, I want to split my data to
training and test datasets. I will use *CARET* for almost every
operation, because it is convenient and smart.

## Precomputing

splitting my data and creating factors:

```{r}
set_seed_for_random_state()

dt_students$gender <- as.factor(dt_students$gender)
dt_students$part_time_job <- as.factor(dt_students$part_time_job)
dt_students$diet_quality <- as.factor(dt_students$diet_quality)
dt_students$parental_education_level <- as.factor(dt_students$parental_education_level)
dt_students$internet_quality <- as.factor(dt_students$internet_quality)
dt_students$extracurricular_participation <- as.factor(dt_students$extracurricular_participation)

dt_train <- createDataPartition(
  y = dt_students$exam_score,
  p = 0.8,
  list = FALSE
)

dt_students_train <- dt_students[dt_train,]
dt_students_test <- dt_students[-dt_train,]
```

## Linear model

So now that we have splitted the data, we can start with Linear model. I
would like to use ten-fold Cross validation, as it from my personal
research is a standard in data science, and repeat it 10 times.

```{r}
set_seed_for_random_state()

train_control <- trainControl(method = "repeatedcv", number = 10, repeats = 10)

lm_scores <- caret::train(exam_score ~ .,
                   data = dt_students_train[, -"student_id"],
                   method = "lm",
                   trControl = train_control,
                   preProcess = c("center","scale"))

summary(lm_scores)
```

R\^2 of 0.9 looks great and now I would like to calculate MSE and then
CV MSE.

Training MSE:

```{r}
set_seed_for_random_state()

scores_lm_train_predictions <- predict(lm_scores, newdata = dt_students_train)

lm_train_mse <- mean((dt_students_train$exam_score - scores_lm_train_predictions)^2)
cat("Training MSE:", lm_train_mse, "\n")
```

<br> We can pull RMSE from CV

```{r}
lm_scores$results
```

<br> calculate MSE:

```{r}
lm_cv_mse <- lm_scores$results$RMSE^2
cat("Cross-Validation MSE:", lm_cv_mse)
```

<br>

and now I want to get MSE from a Test set:

```{r}
scores_lm_test_predictions <- predict(lm_scores, newdata = dt_students_test)

lm_test_mse <- mean((dt_students_test$exam_score - scores_lm_test_predictions)^2)
cat("Test MSE:", lm_test_mse, "\n")
```

## Tree Model

I think this model could perform better, because there some variables
which can affect this model not only linearly, but the other way, in
this case tree model can show better performance.

```{r}
set_seed_for_random_state()

dt_students_train_copy <- copy(dt_students_train)
X_train <- as.data.frame(dt_students_train_copy[, -c("student_id", "exam_score")])
Y_train <- dt_students_train_copy$exam_score

train_control <- trainControl(method = "repeatedcv", 
                              number = 10, 
                              repeats = 10
                              )

# I will let caret to test 20 different cp values
scores_tree <- caret::train(
  x = X_train,
  y = Y_train,
  method = "rpart",
  trControl = train_control,
  tuneLength = 20
)

# print the model
print(scores_tree)
```

I want also find out which cp values were used, and which is chosen

```{r}
best_cp <- scores_tree$bestTune$cp
cat("Best CP:", best_cp, "\n")
```

```{r}
final_scores_tree <- scores_tree$finalModel

prp(final_scores_tree, type = 2, extra = 100, fallen.leaves = TRUE, main = "Final Pruned Decision Tree")
```

One great thing about caret is that it prunes the tree itself, only one
thing I need to do is to set how many different *cp* should caret try.
So now I can calculate MSE for training, CV and test dataset.

So now I can calculate Trainig and CV loss:

```{r}
# training MSE
train_predictions_tree <- predict(scores_tree, newdata = dt_students_train[,-"student_id"])

# Calculate Training MSE
tree_train_mse <- mean((dt_students_train$exam_score - train_predictions_tree)^2)
cat("Tree Model Training MSE:", tree_train_mse, "\n")
```

```{r}
# Extract the RMSE from that row and square it to get MSE
cv_rmse_tree <- scores_tree$results$RMSE[scores_tree$results$cp == best_cp]
tree_cv_mse <- cv_rmse_tree^2
cat("Tree Model Cross-Validation MSE:", tree_cv_mse, "\n")
```

And let's calculate MSE on a test set:

```{r}
# test MSE
test_predictions_tree <- predict(scores_tree, newdata = dt_students_test[, -"student_id"])

# calculate testing MSE
tree_test_mse <- mean((dt_students_test$exam_score - test_predictions_tree)^2)
cat("Tree Model Testing MSE:", tree_test_mse, "\n")
```

## Comparing

So after training and testing linear and decisions tree models, it is
time to compare their performance.

### Linear model performance

```{r}
print_results(lm_train_mse, lm_cv_mse, lm_test_mse)
```

<br> <br>

### Tree model performance

```{r}
print_results(tree_train_mse, tree_cv_mse, tree_train_mse)
```

And I would like to create a small plot to represent visually the difference in performance:

```{r}
values <- c(lm_test_mse, tree_test_mse)
custom_labels <- c("Linear model", "Tree model")

data_for_plot <- data.frame(
  ID = as.factor(custom_labels),
  Value = values
)

ggplot(data_for_plot, aes(x = ID, y = Value)) +
  geom_col(fill = "#69db7c") +
  labs(
    title = "Test Loss of 4 models",
    x = "Model name",
    y = "Loss"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

The linear model performs better than the tree model, and I suppose this
is because my dataset is synthetic and was created for a regression
problem. It is possible, therefore, that the linear model will perform
best overall.

# Ensemble

Now it is time to upgrade tree to a random forest. I will be using caret
for model creation and evaluation.

## Preparation

Before I start with random forest, I need to prepare my hyperparametrs
before training a random forest. There are 2 main hyperparameters:

-   ntree, number of trees grown
-   mtry, number of variables used

```{r}
# preparing the grid for caret
rf_grid <- data.table("mtry"=1:(ncol(dt_students_train)-1))
```

## Training

Now, as I have my hyperparameters, it is time to try it in caret.

```{r}
set_seed_for_random_state()

# control settings
train_control_rf <- trainControl(method = "cv", number = 10)

rf_scores <- caret::train(exam_score ~ ., data = dt_students_train[, -"student_id"],
                   method="rf",
                   ntrees=500,
                   trControl = train_control_rf,
                   search = "grid",
                   tuneGrid = rf_grid,
                   preProcess = c("center","scale"))

rf_scores
```

## Evaluation

Okay, the best mtry value is 13, so I will use this model in my course
work. Now I would like to calculate training MSE

```{r}
# calculating training MSE
train_predictions_rf <- predict(rf_scores, newdata = dt_students_train[, -"student_id"])

rf_train_mse <- mean((dt_students_train$exam_score - train_predictions_rf)^2)

cat("Random Forest Training MSE:", rf_train_mse)
```

```{r}
# Cross Validation MSE
optimal_rf <- rf_scores$results[rf_scores$results$mtry == rf_scores$bestTune$mtry, ]
rf_cv_mse <- optimal_rf$RMSE^2
print(paste("Cross-validation MSE for the optimal mtry (", rf_scores$bestTune$mtry, "):", rf_cv_mse))
```

```{r}
# calculating test MSE
test_predictions_rf <- predict(rf_scores, newdata = dt_students_test)

rf_test_mse <- mean((dt_students_test$exam_score - test_predictions_rf)^2)

cat("Random Forest Test MSE:", rf_test_mse)
```

So, after I trained 500 trees with hyperparameter mtry = 13 I got Train
MSE ~6, which is so much better than a decision tree, but the CV value is ~38.

# Neural Network

I found, that there is Keras3 for R, which is updated version of Keras.
So now I would like to start with Inputs for my NN.

## Preparation

Also it is important to normalize the data, so I will normalize the
data, but from my research, Keras accepts (or TenserFlow under the hood
only numeric variables, so all my factors, should be converted to
numeric values, I will use one-hot encodeing for my factor variables)

```{r}
nn_recipe <- recipe(exam_score ~ ., data = dt_students_train) %>%
  update_role(student_id, new_role = "ID") %>% 
  step_normalize(all_numeric_predictors()) %>% 
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

prepped_recipe <- prep(nn_recipe, training = dt_students_train)
X_train_processed <- bake(prepped_recipe, new_data = dt_students_train)

X_nn_train_matrix <- as.matrix(X_train_processed %>% select(-exam_score, -student_id))
y_nn_train <- dt_students_train$exam_score

X_test_processed <- bake(prepped_recipe, new_data = dt_students_test)
X_nn_test_matrix <- as.matrix(X_test_processed %>% select(-exam_score, -student_id))
y_nn_test <- dt_students_test$exam_score
```

## Configuration 1

### Training

Here I want to strat very simple and use only 2 layers, first input will
have 32 neurons and the second one only 1

Now, after all Inputs are prepared and normalized, it is time to try DNN
in R

```{r}
set_seed_for_random_state()

dnn_scores <- keras_model_sequential() %>%
   layer_dense(32, activation = 'relu') %>%
   layer_dense(1)

dnn_scores %>% compile(
  loss = 'MSE',
  optimizer = "adam"
)

summary(dnn_scores)

```

So now model is created, now it is time to fit it. I also will plot the loss with epochs of training for a better visualization:

```{r}
set_seed_for_random_state()

dnn_history <- dnn_scores %>% fit(
  x=X_nn_train_matrix,
  y=y_nn_train,
  verbose = 0,
  epochs = 100,
  validation_split = 0.2
)

plot(dnn_history)
```

### Evaluation

Calculating training MSE:

```{r}
# predicting on training dataset
dnn_train_predictions <- predict(dnn_scores, X_nn_train_matrix)

# calculating MSE
dnn_train_mse <- mean((dnn_train_predictions - dt_students_train$exam_score)^2)

cat("Training MSE of DNN: ", dnn_train_mse)

```

Calculating test MSE:

```{r}
# predicting
dnn_test_predictions <- predict(dnn_scores, X_nn_test_matrix)

# calculating MSE
dnn_test_mse <- mean((dnn_test_predictions - dt_students_test$exam_score)^2)

cat("Testing MSE of DNN: ", dnn_test_mse)
```

Calculating CV loss:

As this is first place, where I will calculate CV loss for NN, I will
create a reusable function, which will help me along:

```{r}
calculate_cv_loss_for_dnn <- function(data, layers = c(32, 1), epochs = 100, verbose = 0, pca = FALSE) {
  set_seed_for_random_state()

  nn_recipe_cv <- recipe(exam_score ~ ., data = data) %>%
    update_role(student_id, new_role = "ID") %>%
    step_normalize(all_numeric_predictors()) %>%
    step_dummy(all_nominal_predictors(), one_hot = TRUE)
  
  if (pca == TRUE)
    nn_recipe_cv_pca <- recipe(exam_score ~ ., data = data) %>%
      update_role(student_id, new_role = "ID") %>%
      step_normalize(all_numeric_predictors()) %>%
      step_pca(all_numeric_predictors(), threshold = 0.95) %>%
      step_dummy(all_nominal_predictors(), one_hot = TRUE)

  folds <- vfold_cv(data, v = 10)
  cv_losses <- c()

  for (fold in folds$splits) {

    train_data_fold <- analysis(fold)
    val_data_fold <- assessment(fold)

    prepped_recipe_fold <- prep(nn_recipe_cv, training = train_data_fold)

    X_train_processed_fold <- bake(prepped_recipe_fold, new_data = train_data_fold)
    X_val_processed_fold <- bake(prepped_recipe_fold, new_data = val_data_fold)

    X_train_nn_fold <- as.matrix(X_train_processed_fold %>% select(-exam_score, -student_id))
    y_train_fold <- train_data_fold$exam_score
    X_val_nn_fold <- as.matrix(X_val_processed_fold %>% select(-exam_score, -student_id))
    y_val_fold <- val_data_fold$exam_score

    dnn_model <- keras_model_sequential()
    for (i in seq_along(layers)) {
      units <- layers[i]
      if (i == length(layers)) {
        dnn_model <- dnn_model %>% layer_dense(units = units)
      } else {
        dnn_model <- dnn_model %>% layer_dense(units = units, activation = "relu")
      }
    }

    dnn_model %>% compile(loss = "MSE", optimizer = "adam")

    dnn_model %>% fit(
      x = X_train_nn_fold,
      y = y_train_fold,
      epochs = epochs,
      verbose = verbose
    )

    preds <- predict(dnn_model, X_val_nn_fold)
    fold_mse <- mean((preds - y_val_fold)^2)
    cv_losses <- c(cv_losses, fold_mse)
  }

  mean(cv_losses)
}
```

In this function, I pass original data, then inside this data will be
scaled, centered, and then I will use 10 folds to cross validate.

Now, I will calculate CV loss for the configuration with 2 Layers:

```{r}
dnn_cv_mse_1 <- calculate_cv_loss_for_dnn(dt_students_train, c(32, 1))
dnn_cv_mse_1
```

NN with only 2 layers performs better than a decision tree model. Decision Tree has a CV loss of approx 76 and small NN has a test loss of a approx. 28.39.

## Configuration 2

After evaluation of a simple configuration 1, I can try some more complex configurations.

### Training

Here I want to use 3 layers, first input will have
32 neurons, second 16, and the second one only 1

```{r}
set_seed_for_random_state()

dnn_scores_2 <- keras_model_sequential() %>%
   layer_dense(32, activation = 'relu') %>%
   layer_dense(16, activation = 'relu') %>%
   layer_dense(1)

dnn_scores_2 %>% compile(
  loss = 'MSE',
  optimizer = "adam"
)

summary(dnn_scores)

```
fitting:

```{r}
set_seed_for_random_state()

dnn_history_2 <- dnn_scores_2 %>% fit(
  x=X_nn_train_matrix,
  y=y_nn_train,
  verbose = 0,
  epochs = 100,
  validation_split = 0.2
)

plot(dnn_history_2)
```

We can see that with more neurons, model trains faster, and needs less repetitions to get to the optimal parameters.

### Evaluation

Calculating training MSE:

```{r}
# predicting on training dataset
dnn_2_train_predictions <- predict(dnn_scores_2, X_nn_train_matrix)

# calculating MSE
dnn_train_mse_2 <- mean((dnn_2_train_predictions - dt_students_train$exam_score)^2)

cat("Training MSE of DNN: ", dnn_train_mse_2)

```

Calculating test MSE:

```{r}
# predicting
dnn_test_predictions_2 <- predict(dnn_scores_2, X_nn_test_matrix)

# calculating MSE
dnn_test_mse_2 <- mean((dnn_test_predictions_2 - dt_students_test$exam_score)^2)

cat("Testing MSE of DNN: ", dnn_test_mse_2)
```

Calculating CV loss:

```{r}
dnn_cv_mse_2 <- calculate_cv_loss_for_dnn(dt_students_train, c(32, 16, 1))
dnn_cv_mse_2
```
This configuration performs better on training dataset with MSE of ~24.9, but CV loss is almost the same, and simpler model shows slightly better performance.

## Configuration 3

In the last NN configuration, I want to use 4 layers, and I could assume, that this model could overfitt my training dataset and generalize even worser than previous configuration, as second configuration had alredy shown slight generalization decrease.

### Training

Here I want to use 4 layers, 64, 32, 16 and 1

```{r}
set_seed_for_random_state()
dnn_scores_3 <- keras_model_sequential() %>%
  layer_dense(64, activation = 'relu') %>%
  layer_dense(32, activation = 'relu') %>%
  layer_dense(16, activation = 'relu') %>%
  layer_dense(1)

dnn_scores_3 %>% compile(
  loss = 'MSE',
  optimizer = "adam"
)

summary(dnn_scores_3)

```

So now model is created, now I will fit it.

```{r}
set_seed_for_random_state()

dnn_history_3 <- dnn_scores_3 %>% fit(
  x=X_nn_train_matrix,
  y=y_nn_train,
  verbose = 0,
  epochs = 100,
  validation_split = 0.2
)

plot(dnn_history_3)
```

### Evaluating

Calculating training MSE:

```{r}
# predicting on training dataset
dnn_train_predictions_3 <- predict(dnn_scores_3, X_nn_train_matrix)

# calculating MSE
dnn_train_mse_3 <- mean((dnn_train_predictions_3 - dt_students_train$exam_score)^2)

cat("Training MSE of DNN: ", dnn_train_mse_3)

```

Calculating test MSE:

```{r}
# predicting
dnn_test_predictions_3 <- predict(dnn_scores_3, X_nn_test_matrix)

# calculating MSE
dnn_test_mse_3 <- mean((dnn_test_predictions_3 - dt_students_test$exam_score)^2)

cat("Testing MSE of DNN: ", dnn_test_mse_3)
```

Calculating CV loss:

```{r}
dnn_cv_mse_3 <- calculate_cv_loss_for_dnn(dt_students_train, c(64, 32, 16, 1))
dnn_cv_mse_3
```

As I expected CV loss is bigger now, but by a little margin, so it is not a big problem, but it tends to generalize worse if the complexity of the NN grows.

## Best DNN

So now it is time to choose the DNN model, which performs the best:

```{r}
cat("Training performance:\n", "DNN_1: ", dnn_train_mse, "\n", "DNN_2: ", dnn_train_mse_2, "\n", "DNN_3: ", dnn_train_mse_3, "\n\n",
    "Testing performance:\n", "DNN_1: ", dnn_test_mse, "\n", "DNN_2: ", dnn_test_mse_2, "\n", "DNN_3: ", dnn_test_mse_3,  "\n\n",
    "CV Loss:\n", "DNN_1: ", dnn_cv_mse_1, "\n", "DNN_2: ", dnn_cv_mse_2, "\n", "DNN_3: ", dnn_cv_mse_3)
```

From all models, I think the largest one is slightly overfitting the data
and thus generalizes slightly worse than smallest model in my case.
So I will chose the first model to compare with others. And here we can take a look on it:

```{r}
summary(dnn_scores)
```


# Model Comparison

Let's compare all 4 models, which I have now. There are Linear model, tree, random forest, deep neural network. I want to compare them by CV loss and for that I will make small bar plot with their performance:

```{r}
values <- c(lm_cv_mse, tree_cv_mse, rf_cv_mse, dnn_cv_mse_1)
custom_labels <- c("Linear model", "Tree model", "Random forest", "DNN")

data_for_plot <- data.frame(
  ID = as.factor(custom_labels),
  Value = values
)

ggplot(data_for_plot, aes(x = ID, y = Value)) +
  geom_col(fill = "#69db7c") +
  geom_text(aes(label = round(Value, 2)), 
            vjust = -0.5,                  
            size = 3) +                    
  labs(
    title = "Test Loss of Various Models", 
    x = "Model name",
    y = "Loss (MSE)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 7),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")
  )
```
Among the models evaluated, the linear model delivered the best performance, with DNN and Random Forest models yielding very comparable results. Notably, the decision tree model exhibited significantly higher errors, nearly tripple that of the linear model. This outcome is intriguing and suggests the underlying true function within this synthetic dataset might indeed be linear. But maybe DNN and Random forest will provide better results after performing PCA.

One more thing, that is also interesting for me is training loss:

```{r}
values <- c(lm_train_mse, tree_train_mse, rf_train_mse, dnn_train_mse)
custom_labels <- c("Linear model", "Tree model", "Random forest", "DNN")

data_for_plot <- data.frame(
  ID = as.factor(custom_labels),
  Value = values
)

ggplot(data_for_plot, aes(x = ID, y = Value)) +
  geom_col(fill = "#69db7c") +
  geom_text(aes(label = round(Value, 2)), 
            vjust = -0.5,                  
            size = 3) +                    
  labs(
    title = "Test Loss of Various Models", 
    x = "Model name",
    y = "Loss (MSE)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 7),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")
  )
```

The behavior of the Random Forest model is particularly surprising. While Random Forests are typically employed to mitigate variance, it notably exhibited the lowest training loss among all models, and by a significant margin.

# PCA

After very interesting investigation and comparement of 4 models, I would like to run PCA and compare CV loss from the all models again.

## Preparation

As I used before the library recipies from tidymodels, I found it very convenient to work with this library, and also it provides PCA, so I will create new Recipie for a PCA part of the work.

```{r}
nn_pca_recipe <- recipe(exam_score ~ ., data = dt_students_train) %>%
  update_role(student_id, new_role = "ID") %>%
  step_normalize(all_numeric_predictors()) %>% 
  step_pca(all_numeric_predictors(), threshold = 0.95) %>% 
  step_dummy(all_nominal_predictors(), one_hot = TRUE) 

# preparing transformation on a basis of train dataset
prepped_pca_recipe <- prep(nn_pca_recipe, training = dt_students_train)

# applying transfomration to my train and test datasets
X_train_pca_processed <- bake(prepped_pca_recipe, new_data = dt_students_train)
X_test_pca_processed <- bake(prepped_pca_recipe, new_data = dt_students_test)

# preparing X and y values for models
X_train_pca <- as.data.frame(X_train_pca_processed %>% select(-exam_score, -student_id))
y_train_pca <- dt_students_train$exam_score

X_test_pca <- X_test_pca_processed %>% select(-exam_score, -student_id)
y_test_pca <- dt_students_test$exam_score

# For Keras, we often need matrices directly
X_nn_train_pca_matrix <- as.matrix(X_train_pca)
X_nn_test_pca_matrix <- as.matrix(X_test_pca)
```

So now the data is prepared to re run all models on PCs.

## Linear model

```{r}
set_seed_for_random_state()

train_control <- trainControl(method = "repeatedcv", number = 10, repeats = 10)

lm_pca_scores <- caret::train(exam_score ~ .,
                           data = cbind(X_train_pca, exam_score = y_train_pca),
                           method = "lm",
                           trControl = train_control)

summary(lm_pca_scores)

# Training MSE
lm_pca_train_predictions <- predict(lm_pca_scores, newdata = X_train_pca)
lm_pca_train_mse <- mean((y_train_pca - lm_pca_train_predictions)^2)
cat("LM PCA Training MSE:", lm_pca_train_mse, "\n")

# CV MSE
lm_pca_cv_mse <- lm_pca_scores$results$RMSE^2
cat("LM PCA Cross-Validation MSE:", lm_pca_cv_mse, "\n")

# Test MSE
lm_pca_test_predictions <- predict(lm_pca_scores, newdata = X_test_pca)
lm_pca_test_mse <- mean((y_test_pca - lm_pca_test_predictions)^2)
cat("LM PCA Test MSE:", lm_pca_test_mse, "\n")
```


## Tree model

```{r}
set_seed_for_random_state()

train_control <- trainControl(method = "repeatedcv",
                              number = 10,
                              repeats = 10)

scores_tree_pca <- caret::train(
  x = X_train_pca,
  y = y_train_pca,
  method = "rpart",
  trControl = train_control,
  tuneLength = 20
)

print(scores_tree_pca)

# Training MSE
tree_pca_train_predictions <- predict(scores_tree_pca, newdata = X_train_pca)
tree_pca_train_mse <- mean((y_train_pca - tree_pca_train_predictions)^2)
cat("Tree PCA Training MSE:", tree_pca_train_mse, "\n")

# CV MSE
best_cp_pca <- scores_tree_pca$bestTune$cp
cv_rmse_tree_pca <- scores_tree_pca$results$RMSE[scores_tree_pca$results$cp == best_cp_pca]
tree_pca_cv_mse <- cv_rmse_tree_pca^2
cat("Tree PCA Cross-Validation MSE:", tree_pca_cv_mse, "\n")

# Test MSE
test_predictions_tree_pca <- predict(scores_tree_pca, newdata = X_test_pca)
tree_pca_test_mse <- mean((y_test_pca - test_predictions_tree_pca)^2)
cat("Tree PCA Test MSE:", tree_pca_test_mse, "\n")
```

## Random forest

```{r}
set_seed_for_random_state()

rf_pca_grid <- data.table("mtry" = 1:ncol(X_train_pca))

train_control_rf <- trainControl(method = "cv", number = 10)

rf_pca_scores <- caret::train(x = X_train_pca,
                           y = y_train_pca,
                           method = "rf",
                           ntrees = 500,
                           trControl = train_control_rf,
                           search = "grid",
                           tuneGrid = rf_pca_grid)

rf_pca_scores

# Training MSE
train_predictions_rf_pca <- predict(rf_pca_scores, newdata = X_train_pca)
rf_pca_train_mse <- mean((y_train_pca - train_predictions_rf_pca)^2)
cat("RF PCA Training MSE:", rf_pca_train_mse, "\n")

# Cross Validation MSE
optimal_rf_pca <- rf_pca_scores$results[rf_pca_scores$results$mtry == rf_pca_scores$bestTune$mtry, ]
rf_pca_cv_mse <- optimal_rf_pca$RMSE^2
print(paste("RF PCA Cross-validation MSE for optimal mtry (", rf_pca_scores$bestTune$mtry, "):", rf_pca_cv_mse))

# Test MSE
test_predictions_rf_pca <- predict(rf_pca_scores, newdata = X_test_pca)
rf_pca_test_mse <- mean((y_test_pca - test_predictions_rf_pca)^2)
cat("RF PCA Test MSE:", rf_pca_test_mse, "\n")
```

## DNN

```{r}
set_seed_for_random_state()

dnn_scores_pca <- keras_model_sequential() %>%
   layer_dense(32, activation = 'relu') %>%
   layer_dense(1)

dnn_scores_pca %>% compile(
  loss = 'MSE',
  optimizer = "adam"
)

summary(dnn_scores_pca)

dnn_history <- dnn_scores_pca %>% fit(
  x=X_nn_train_pca_matrix,
  y=y_nn_train,
  verbose = 0,
  epochs = 100,
  validation_split = 0.2
)

plot(dnn_history)

```

```{r}
# predicting on training dataset
dnn_train_predictions_pca <- predict(dnn_scores_pca, X_nn_train_pca_matrix)

# calculating MSE
dnn_train_pca_mse <- mean((dnn_train_predictions_pca - dt_students_train$exam_score)^2)

cat("Training MSE of DNN: ", dnn_train_pca_mse)
```

```{r}
# predicting
dnn_test_predictions_pca <- predict(dnn_scores_pca, X_nn_test_pca_matrix)

# calculating MSE
dnn_test_mse_pca <- mean((dnn_test_predictions_pca - dt_students_test$exam_score)^2)

cat("Testing MSE of DNN: ", dnn_test_mse_pca)
```

```{r}
dnn_cv_mse_pca <- calculate_cv_loss_for_dnn(dt_students_train, c(32, 1), 100, 0, TRUE)
dnn_cv_mse_pca
```


# Model Selection

-   Compare all of your models based on their CV performance. You will
    have 8 options to consider: four models with and without PCA.
-   Present the results in a table or chart.
-   Estimate the expected loss of your best model on the test set.


```{r}
values <- c(lm_cv_mse, tree_cv_mse, rf_cv_mse, dnn_cv_mse_1, lm_pca_cv_mse, tree_pca_cv_mse, rf_pca_cv_mse, dnn_cv_mse_pca)
custom_labels <- c("Linear model", "Tree model", "Random forest", "DNN", "Linear model PCA", "Tree model PCA", "Random forest PCA", "DNN PCA")

data_for_plot <- data.frame(
  ID = as.factor(custom_labels),
  Value = values
)

ggplot(data_for_plot, aes(x = ID, y = Value)) +
  geom_col(fill = "#69db7c") +
  geom_text(aes(label = round(Value, 2)), 
            vjust = -0.5,                  
            size = 3) +                    
  labs(
    title = "Test Loss of Various Models", 
    x = "Model name",
    y = "Loss (MSE)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 7),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.margin = unit(c(1, 1, 1, 1), "lines")
  )
```


