---
title: "Coursework - Data Science II"
author: "Omar Zhadykov, 220220503"
output:
  html_notebook:
    fig_width: 10
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document: default
  html_document:
    fig_width: 10
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
---

<script>
$(document).ready(function() {
  $items = $('div#TOC li');
  $items.each(function(idx) {
    num_ul = $(this).parentsUntil('#TOC').length;
    $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
  });

});
</script>

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}
library(svglite)
library(knitr)
suppressPackageStartupMessages(library(data.table))
library(ggplot2)
knitr::opts_chunk$set(dev = "svglite")

# Put your dataset in the same folder as your R file. This code will set your working directory for this notebook to the folder where the R file is stored. This way I can rerun your code without modifications.

#library(rstudioapi)
#setwd(dirname(getActiveDocumentContext()$path))

library(here)
setwd(here())

print_results <- function(train, cv, test) {
  cat("Training MSE: ", train, "\n")
  cat("CV Loss: ", cv, "\n")
  cat("Test MSE: ", test)
}

```

# Introduction

In this course work I will explore dataset about students' performance based on their habits and lifestyle. I have chosen this dataset, because it is rich on different discrete, continious variables with a good range and I assume good variance, which will be very useful in Data Science 2 class. 

This dataset consists of 16 columns which could be categorized into these groups:
  - discrete variables (example: exercise frequency per week)
  - continious variables (example: sleep quantity)
  - categorical variables (example par time job, yes or no)
  
This is regression problem, because I will try to predict student's score for an exam, based on different variables.

### Collection

And before I start, I would like to create a data table where, I will store performance of the models

```{r}
models_performance <- list()
```


Here I would like to collect, prepare, and explore my data. First thing is to import the data set.

```{r}
dt_students <- fread(file = "./data/student_habits_performance.csv")
```

<br>

I would like to check, if i have some nullish data in my dataset. I think it is a good idea to go through all rows and colums and check, if there is a NA. I want to check it with built-in function in R *complete.cases(data_table)*. This function returns TRUE or FALSE if row contains a NA value.

```{r}
nas <- dt_students[!complete.cases(dt_students)]
nas
```

That looks great. Now we can move on to exploration. But before I start, It is crucial to install all needed libraries.

```{r}
library(data.table)
library(ggcorrplot)
library(ggExtra)
library(ggplot2)
library(skimr)
library(ggridges)
library(ggsci)
library(ggthemes)
library(RColorBrewer)
library(caret)
library(svglite)
library(viridis)
library(scales)
library(rpart)
library(rpart.plot)
library(factoextra)
library(keras3)
library(tensorflow)
library(dplyr)
library(reticulate)

# this is for replication
set.seed(42)
tensorflow::tf$random$set_seed(42)
py_run_string("import numpy as np; import random; random.seed(42); np.random.seed(42)")
```

### Exploration

I found some helpful functions in R, so we could have a look on our data. We will start with a structure, than we will get some statistic data and take a *head()* of the data

```{r}
str(dt_students)
```
<br>
Statistic data:
```{r}
skim(dt_students)
```

So we have 1000 rows, 16 columns, 7 char values and 9 numeric values. This is great start to get an overall opinion, so let's move.

<br>
This is a sample of dataset:

```{r}
head(dt_students)
```

I would like to start from density of a main values, which are from my domain knowledge are important for the best performance at the university.

Attendance percentage density: 

```{r}
ggplot(data = dt_students, aes(x = attendance_percentage)) + 
  geom_density(fill="#f1b147", color="#f1b147", alpha=0.25) + 
  labs(
    x = 'Attendance',
    y = 'Density'
  ) +
  geom_vline(xintercept = mean(dt_students$attendance_percentage), linetype="dashed") + 
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```
This density plot visualizes the distribution of the attendance percentage, showing that most students attend classes at a rate of ~85% roughly, and this is a right-skewed distribution. The dashed vertical line represents the mean attendance percentage (~84-85%). The plot shows, that the majority of students are attending most of the classes.


Age density:

```{r}
ggplot(data = dt_students, aes(x = age)) + 
  geom_density(fill="#f1b147", color="#f1b147", alpha=0.25) + 
  labs(
    x = 'Age',
    y = 'Density'
  ) +
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

Age density plot looks pretty well distributed, so I think there is a better chance to learn good model from this dataset and generalize well.


<br>

Next plot will visualize the distribution of exam scores depending on attendance, and I would like to take a look at how possibly hours studied per day affect the exam score. 

```{r}
ggplot() + 
  geom_point(data = dt_students, aes(x = attendance_percentage, y = exam_score, color = study_hours_per_day)) +
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

This scatter plot visualizes the relationship between attendance percentage (x-axis) and exam score (y-axis), with color indicating the amount of hours studied per day. It shows a positive correlation between high attendance, more study hours per day and exam score. I would say that on average students, who learn and attend classes more tend to get better scores. However, there is some variability, as some students with attendance more than 85% and average studying hours get worst scores.

The next plot, which I am going to do is a boxplot and I want to use diet quality as a factor variable on x axis and exam score on y-axis, to get an overall understanding, how the diet quality may affect the exam score.

```{r}
ggplot(data = dt_students, aes(x = factor(diet_quality), y = exam_score)) +
  geom_boxplot() + 
  theme_minimal() 
```

Boxplot shows that diet quality could be not a very valuable predictor variable. But, we can see, that poor diet has a little bit lower performance. From my personal opinion this difference is so statistically low, that it will not play a big role, but this is still interesting, that If person wants to get the best performance, it is also important to get "the best fuel".

Another very important thing is sleep. I would like to know if there is visible relation between sleep, exercise frequency and exam score, and also sleep distribution of the students.

```{r}
ggplot(data = dt_students, aes(x = sleep_hours, y = exam_score, colour = exercise_frequency)) +
  geom_point() +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000")) +
  scale_colour_gradient(low = "green", high = "darkgreen")
```

```{r}
ggplot(data = dt_students, aes(x = sleep_hours)) + 
  geom_density(fill="#f1b147", color="#f1b147", alpha=0.25) + 
  labs(
    x = 'Sleep',
    y = 'Density'
  ) +
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```

From this two plots it is clear that it is not very clear that sleep and sport play a very important role, because there are students who sleep, does sports and gets great scores, but also students, who does not sleep as much, and doesn't so much of a sport, but still gets good grades. Even some Students with 4.5 hours and 1 workaout a week are getting 100% exam score.


Let's create one more boxplot to look at average notes of students with and without part time jobs:

```{r}
ggplot(data = dt_students, aes(x = factor(part_time_job), y = exam_score)) +
  geom_boxplot() + 
  theme_minimal() 
```

Boxplot shows some little difference, but as i thought it is roughly the same.


Now I would like to take a look at one more scatter plot with amount of social media hours with exam score and gender.

```{r}
ggplot( data = dt_students, aes(x = social_media_hours, y = exam_score, colour = gender)) +
  geom_point() +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000"))
```

I would say that I expected another result, but it seems, that on average amount of social media usage does not play a very big role, but it is also an slight average better to spend less time on social media to get better scores.

One last plot will include mental health rating, which from my point of view is very important to perform well.

```{r}
ggplot(data = dt_students, aes(x = mental_health_rating, y = exam_score, colour = gender)) +
  geom_point() +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000"))
```

This plot does not provide enough information, so let's take a look on a vertical barchart.

```{r}
ggplot(data = dt_students, aes(x = mental_health_rating)) + 
  geom_histogram(fill="#2f9e44", color="#2f9e44", alpha=0.25) + 
  geom_vline(xintercept = mean(dt_students$mental_health_rating), linetype="dashed") + 
  theme_minimal() + 
  theme(axis.line = element_line(color = "#000000"))
```


```{r}
mean(dt_students$mental_health_rating)
```

Okay, from the last 2 plots I could conclude, that mental health could possibly better results on average, but students can still perform well, while being in a "bad" mood.


I think that would be enough exploration and we can start with models.


# Models 1 & 2

Now the interesting part! So before I start, I want to split my data to training and test datasets. I will use *CARET* for almost every operations, because it is convinient and smart.

## Precomputing

splitting my data and creating factors:


```{r}
dt_students$gender <- as.factor(dt_students$gender)
dt_students$part_time_job <- as.factor(dt_students$part_time_job)
dt_students$diet_quality <- as.factor(dt_students$diet_quality)
dt_students$parental_education_level <- as.factor(dt_students$parental_education_level)
dt_students$internet_quality <- as.factor(dt_students$internet_quality)
dt_students$extracurricular_participation <- as.factor(dt_students$extracurricular_participation)

dt_train <- createDataPartition(
  y = dt_students$exam_score,
  p = 0.8,
  list = FALSE
)

dt_students_train <- dt_students[dt_train,]
dt_students_test <- dt_students[-dt_train,]
```


## Linear model

So now that we have splitted the data, we can start with Linear model. I would like to use ten-fold Cross validatian, as it from my personal
research is a standard in data science, and repeat it 10 times.

```{r}
train_control <- trainControl(method = "repeatedcv", number = 10, repeats = 10)

lm_scores <- caret::train(exam_score ~ .,
                   data = dt_students[, -"student_id"],
                   method = "lm",
                   trControl = train_control,
                   preProcess = c("center","scale"))

summary(lm_scores)
```

R^2 of 0.9 looks great and now I would like to calculate MSE and then CV MSE.

Training MSE:
```{r}
scores_lm_train_predictions <- predict(lm_scores, newdata = dt_students_train)

lm_train_mse <- mean((dt_students_train$exam_score - scores_lm_train_predictions)^2)
cat("Training MSE:", lm_train_mse, "\n")
```

<br>
We can pull RMSE from CV

```{r}
lm_scores$results
```
<br>
calculate MSE:
```{r}
lm_cv_mse <- lm_scores$results$RMSE^2
cat("Cross-Validation MSE:", lm_cv_mse)
```
<br>

and now I want to get MSE from a Test set:
```{r}
scores_lm_test_predictions <- predict(lm_scores, newdata = dt_students_test)

lm_test_mse <- mean((dt_students_test$exam_score - scores_lm_test_predictions)^2)
cat("Test MSE:", lm_test_mse, "\n")
```


## Tree Model

I think this model could perform better, because there some variables which can affect this model not only linearly, but the other way, in this case tree model can show better performance.

```{r}
dt_students_train_copy <- copy(dt_students_train)
X_train <- as.data.frame(dt_students_train_copy[, -c("student_id", "exam_score")])
Y_train <- dt_students_train_copy$exam_score

train_control <- trainControl(method = "repeatedcv", 
                              number = 10, 
                              repeats = 10,
                              )

# I will let caret to test 20 different cp values
scores_tree <- caret::train(
  x = X_train,
  y = Y_train,
  method = "rpart",
  trControl = train_control,
  tuneLength = 20,
  preProcess = c("center", "scale")
)

# print the model
print(scores_tree)
```

I want also find out which cp values were used, and which is chosen

```{r}
best_cp <- scores_tree$bestTune$cp
cat("Best CP:", best_cp, "\n")
```

```{r}
final_scores_tree <- scores_tree$finalModel

prp(final_scores_tree, type = 2, extra = 100, fallen.leaves = TRUE, main = "Final Pruned Decision Tree")
```

One great thing about caret is that it prunes the tree itself, only one thing I need to do is to set how many different *cp* should caret try. So now I can calculate MSE for training, CV and test dataset.


So now I can calculate Trainig and CV loss:

```{r}
# training MSE
train_predictions_tree <- predict(scores_tree, newdata = dt_students_train[,-"student_id"])

# Calculate Training MSE
tree_train_mse <- mean((dt_students_train$exam_score - train_predictions_tree)^2)
cat("Tree Model Training MSE:", tree_train_mse, "\n")
```

```{r}
# Extract the RMSE from that row and square it to get MSE
cv_rmse_tree <- scores_tree$results$RMSE[scores_tree$results$cp == best_cp]
tree_cv_mse <- cv_rmse_tree^2
cat("Tree Model Cross-Validation MSE:", tree_cv_mse, "\n")
```

And let's calculate MSE on a test set:

```{r}
# test MSE
test_predictions_tree <- predict(scores_tree, newdata = dt_students_test[, -"student_id"])

# calculate testing MSE
tree_test_mse <- mean((dt_students_test$exam_score - test_predictions_tree)^2)
cat("Tree Model Testing MSE:", tree_test_mse, "\n")
```

## Comparing

So after training and testing linear and decisions tree models, it is time to compare their performance.

### Linear model performance
```{r}
print_results(lm_train_mse, lm_cv_mse, lm_test_mse)
```
<br>
<br>

### Tree model performance
```{r}
print_results(tree_train_mse, tree_cv_mse, tree_train_mse)
```

The linear model performs better than the tree model, and I suppose this is because my dataset is synthetic and was created for a regression problem. It is possible, therefore, that the linear model will perform best overall.

# Ensemble 

Now it is time to upgrade tree to a random forest. I will be using caret for model creation and eveluation.

## Preparation

Before I start with random forest, I need to prepare my hyperparametrs before training a random forest. There are 2 main hyperparameters: 

 - ntree, number of trees grown
 - mtry, number of variables used

```{r}
# preparing the grid for caret
rf_grid <- data.table("mtry"=1:(ncol(dt_students_train)-1))
```

Now, as I have my hyperparameters, it is time to try it in caret.

```{r}
# controll settings
train_control_rf <- trainControl(method = "cv", number = 10)

rf_scores <- caret::train(exam_score ~ ., data = dt_students_train[, -"student_id"],
                   method="rf",
                   ntrees=500,
                   trControl = train_control_rf,
                   search = "grid",
                   tuneGrid = rf_grid)

rf_scores
```

Okay, the best mtry value is 11, so I will use this model in my course work. Now I would like to calculate training MSE

```{r}
# calculating training MSE
train_predictions_rf <- predict(rf_scores, newdata = dt_students_train[, -"student_id"])

rf_train_mse <- mean((dt_students_train$exam_score - train_predictions_rf)^2)

cat("Random Forest Training MSE:", rf_train_mse)
```

```{r}
# Cross Validation MSE
optimal_rf <- rf_scores$results[rf_scores$results$mtry == rf_scores$bestTune$mtry, ]
rf_cv_mse <- optimal_rf$RMSE^2
print(paste("Cross-validation MSE for the optimal mtry (", rf_scores$bestTune$mtry, "):", rf_cv_mse))
```

```{r}
# calculating test MSE
test_predictions_rf <- predict(rf_scores, newdata = dt_students_test)

rf_test_mse <- mean((dt_students_test$exam_score - test_predictions_rf)^2)

cat("Random Forest Test MSE:", rf_test_mse)
```


So, after I trained 500 trees with hyperparameter mtry = 11 I got Test MSE 46.08, which is 40% better than regular tree. Next is NN up. 

# Neural Network

- Repeat the analysis with a neural network. 
- Investigate three different configurations with regards to your evaluation metric and select the best configuration. Use the same evaluation metric as before.
- Report both the training and the CV loss. 

I found, that there is Keras3 for R, which is updated version of Keras. So now I would like to start with Inputs for my NN.

```{r}
y_nn <- dt_students_train$exam_score
X_df <- dt_students_train[,-c("student_id", "exam_score")]
```

Also it is important to normalize the data, so I will normalize the data, but from my research, Keras accepts (or TenserFlow under the hood only numeric variables, so all my factors, should be converted to numeric values, I will use one-hot encodeing for my factor variables)

```{r}
# scaling numeric 
X_df_scaled <- X_df %>%
  mutate_if(is.numeric, scale)

# one-hot encoding for factors
X_nn <- model.matrix(~ . -1, data = X_df_scaled)

```


## Configuration 1

Here I want to strat very simple and use only 2 layers, first input will have 32 neurons and the second one only 1

Now, after all Inputs are prepared and normilized, it is time to try DNN in R

```{r}
dnn_scores <- keras_model_sequential() %>%
   layer_dense(32, activation = 'relu') %>%
   layer_dense(1)

dnn_scores %>% compile(
  loss = 'MSE',
  optimizer = "adam"
)

summary(dnn_scores)

```

So now model is created, now it is time to fit it.

```{r}
dnn_history <- dnn_scores %>% fit(
  x=X_nn,
  y=y_nn,
  verbose = 0,
  epochs = 100,
  validation_split = 0.2
)

plot(dnn_history)
```

And collecting the results

```{r}
models_performance[['dnn_scores']] <- dnn_scores %>% evaluate(
  X_nn,
  y_nn,
  verbose = 0
)
```


Calculating training MSE:

```{r}
# predicting on training dataset
dnn_train_predictions <- predict(dnn_scores, X_nn)

# calculating MSE
dnn_train_mse <- mean((dnn_train_predictions - dt_students_train$exam_score)^2)

cat("Training MSE of DNN: ", dnn_train_mse)

```


Calculating test MSE:

```{r}
# first I need to prepare train dataset before passing it to DNN model

# slicing useless cols
X_test_df <- dt_students_test[,-c("student_id", "exam_score")]

# scaling numeric 
X_test_df_scaled <- X_test_df %>%
  mutate_if(is.numeric, scale)

# one-hot encoding for factors
X_test_nn <- model.matrix(~ . -1, data = X_test_df_scaled)

# predicting
dnn_test_predictions <- predict(dnn_scores, X_test_nn)

# calculating MSE
dnn_test_mse <- mean((dnn_test_predictions - dt_students_test$exam_score)^2)

cat("Testing MSE of DNN: ", dnn_test_mse)
```

## Configuration 2

Here I want to strat very simple and use 3 layers, first input will have 32 neurons, second 16, and the second one only 1

```{r}
dnn_scores_2 <- keras_model_sequential() %>%
   layer_dense(32, activation = 'relu') %>%
   layer_dense(16, activation = 'relu') %>%
   layer_dense(1)

dnn_scores_2 %>% compile(
  loss = 'MSE',
  optimizer = "adam"
)

summary(dnn_scores)

```

fitting:

```{r}
dnn_history_2 <- dnn_scores_2 %>% fit(
  x=X_nn,
  y=y_nn,
  verbose = 0,
  epochs = 100,
  validation_split = 0.2
)

plot(dnn_history_2)
```

And collecting the results

```{r}
models_performance[['dnn_scores_2']] <- dnn_scores_2 %>% evaluate(
  X_nn,
  y_nn,
  verbose = 0
)
```


Calculating training MSE:

```{r}
# predicting on training dataset
dnn_2_train_predictions <- predict(dnn_scores_2, X_nn)

# calculating MSE
dnn_train_mse_2 <- mean((dnn_2_train_predictions - dt_students_train$exam_score)^2)

cat("Training MSE of DNN: ", dnn_train_mse_2)

```

Calculating test MSE:

```{r}
# first I need to prepare train dataset before passing it to DNN model

# slicing useless cols
X_test_df <- dt_students_test[,-c("student_id", "exam_score")]

# scaling numeric 
X_test_df_scaled <- X_test_df %>%
  mutate_if(is.numeric, scale)

# one-hot encoding for factors
X_test_nn <- model.matrix(~ . -1, data = X_test_df_scaled)

# predicting
dnn_test_predictions_2 <- predict(dnn_scores_2, X_test_nn)

# calculating MSE
dnn_test_mse_2 <- mean((dnn_test_predictions_2 - dt_students_test$exam_score)^2)

cat("Testing MSE of DNN: ", dnn_test_mse_2)
```


## Configuration 3

Here I want to use 4 layers, 64, 32, 16 and 1

```{r}
dnn_scores_3 <- keras_model_sequential() %>%
  layer_dense(64, activation = 'relu') %>%
  layer_dense(32, activation = 'relu') %>%
  layer_dense(16, activation = 'relu') %>%
  layer_dense(1)

dnn_scores_3 %>% compile(
  loss = 'MSE',
  optimizer = "adam"
)

summary(dnn_scores_3)

```

So now model is created, now I will fit it.

```{r}
dnn_history_3 <- dnn_scores_3 %>% fit(
  x=X_nn,
  y=y_nn,
  verbose = 0,
  epochs = 100,
  validation_split = 0.2
)

plot(dnn_history_3)
```

And collecting the results

```{r}
models_performance[['dnn_scores_3']] <- dnn_scores_3 %>% evaluate(
  X_nn,
  y_nn,
  verbose = 0
)
```


Calculating training MSE:

```{r}
# predicting on training dataset
dnn_train_predictions_3 <- predict(dnn_scores_3, X_nn)

# calculating MSE
dnn_train_mse_3 <- mean((dnn_train_predictions_3 - dt_students_train$exam_score)^2)

cat("Training MSE of DNN: ", dnn_train_mse_3)

```


Calculating test MSE:

```{r}
# first I need to prepare train dataset before passing it to DNN model

# slicing useless cols
X_test_df <- dt_students_test[,-c("student_id", "exam_score")]

# scaling numeric 
X_test_df_scaled <- X_test_df %>%
  mutate_if(is.numeric, scale)

# one-hot encoding for factors
X_test_nn <- model.matrix(~ . -1, data = X_test_df_scaled)

# predicting
dnn_test_predictions_3 <- predict(dnn_scores_3, X_test_nn)

# calculating MSE
dnn_test_mse_3 <- mean((dnn_test_predictions_3 - dt_students_test$exam_score)^2)

cat("Testing MSE of DNN: ", dnn_test_mse_3)
```

## Best DNN

So now it is time to choose the DNN model, which performs the best:

```{r}
cat("Training performance:\n", "DNN_1: ", dnn_train_mse, "\n", "DNN_2: ", dnn_train_mse_2, "\n", "DNN_3: ", dnn_train_mse_3, "\n\n")
cat("Testing performance:\n", "DNN_1: ", dnn_test_mse, "\n", "DNN_2: ", dnn_test_mse_2, "\n", "DNN_3: ", dnn_test_mse_3)
```

From all models, I think the largest one is slightly overfitting data and thus generalizes slightly worse than middle sized model in my case. So I will chose the second model to compare with others.


# Model Comparison

- Compare the performance of the 4 models.

# PCA

- Run a PCA on your input variables and discuss the scope for dimensionality reduction in your dataset.
- Rerun the previous 4 models on all PCs or on a reduced number of PCs.

### Calculating PCs

Before I start working with PCA, this is important to normalize the data, so that measurement scale will not affect PCs.

```{r}
#dt_pca <- data.table(scale(dt_houses[,c("area", "bedrooms", "bathrooms", "stories", "parking")]))
```

Now I want to run my PCA with help of prcomp and get the summary, to dive in to Data.

```{r}
#dt_houses_pca <- prcomp(dt_pca)
#summary(dt_houses_pca)
```
That looks interesting, but I would like to plot it, to visualize it, so that it could be easier to understand.

```{r}
#fviz_eig(dt_houses_pca, addlabels = TRUE) + 
#         theme(plot.title = element_text(hjust = 0.5))
```

From this plot it is obvious that all PC are useful and reducing dimensionality will not be beneficial, because last 3 PCs contribute approx. 12% each, which is a fair amount in this case. So I would like to use all PCs, because it is not always necessary to cut dimensionality and it depends on how much variablse we actually have, and how big of a contribution makes each PC. May be in this case "change of basis" will perform better.

### Running models with PCs

Now as I have calculated PCs, it is time to run models with new inputs.

#### Linear model

First we will start off with linear model.

```{r}
# pc_table <- dt_houses_pca$x

# dt_houses_pc_table_for_lm <- data.frame(price = dt_houses$price, pc_table)

# price_lm_pca <- lm(formula = price ~ ., data = dt_houses_pc_table_for_lm)

# summary(price_lm_pca)
```

```{r}
# price_lm_pca_rmse <- mean(sqrt(abs(price_lm_pca$residuals)))

# price_lm_pca_rmse
```


Now, after PCA the results are worse and I think this is because there are fair amount of binary variables in this dataset and PCA could not capture all information, espesially information, which is captured in binary variables, such as "mainroad", "airconditioning" and so on.

 
# Model Selection

- Compare all of your models based on their CV performance. You will have 8 options to consider: four models with and without PCA.
- Present the results in a table or chart.
- Estimate the expected loss of your best model on the test set.

***